Methods

### 2021-10-05 Manuscript/Manuscript_Methods.Rmd
various edits

A first initial set of manually labelled records is needed to start the automatic classification. We suggest manually label the first 250 records (see "hyperparameter optimization" later).\

### Document-Term matrix generation
The annotation file produced in the previous step has a number of fields characterizing a scientific publication. The framework models the relevance of a record based on the following fields: title, abstract, authors, keywords, MESH terms [@lipscomb2000medical]. A series of Natural Language Processing (NLP) techniques [@marshall2015systematic; @ananiadou2006text] are employed to transform the textual information in these fields into features for machine learning, through a bag-of-words approach [@marshall2015systematic]. The processing of free text fields includes tokenization (i.e., extracting the terms), common stopwords (i.e. sentence components bringing no meaning) removal, part-of-speech filtering (only nouns, adjectives, verbs and untagged terms are retained), and lemmatization of the terms (i.e. reduction to their base grammar form). Processing for authors, keywords and MESH terms identify logical units (e.g., author's full names, composite keywords) without breaking them into single terms, and no stopwords removal or lemmatization are applied.\
To reduce noise and save computation time, terms that appeared in less than 5% of the labelled documents (positive and negatives) are removed from negative records. All terms in the positive set are kept to increase sensitivity at the cost of specificity.\
Some terms tend to co-appear in records (non-consecutive ngrams, nc-ngrams), often carrying a particular meaning when copresent. To detect nc-ngrams, we generated a network representation posing edges between terms with a cosine similarity in terms of records copresence > 0.5. We extracted the maximal cliques in the network [@eppstein2010listing] representing highly correlated groups of terms; These generated terms are added to the data set. To avoid overfitting, we kept nc-ngrams of a maximum of ten terms.\
A similarity network is built again using a similarity threshold of 0.9 and finding the cliques again. In this case, the cliques represent terms that always appear together and therefore can be considered synonyms. These terms are merged in the data set to increase computation efficiency and reduce overfitting.\
The final output is a matrix, also called a Document-Term Matrix (DTM), with $N_d$ rows representing the records $D$, $N_t$ terms column for the $t_{field}$ terms (divided by record field) and ${0,1}$ values whether $t_{field} \in D$. We also enriched the DTM with features storing the number of terms in each field to help the model scale term importance based on the document length.

@@

$$\pi_i = \frac{1}{M}\sum_mPr(L_i = \text{pos}|DTM,model_m)$$

as the PPD of a record $D_i$ being assigned a positive label ($L_i$), averaging the PPDs of the ensemble of $M=10$ models, and

@@

* We fit a partition tree [@rpart] between the DTM and 800 samples from the PPD; if a term is present multiple times in the DTM (e.g. both title and abstract), they are counted just one, and field term count features are removed. This step generates a list of rules composed by $AND$/$NOT$ "conditions" made of terms/authors/keywords/MESH, which together identify a group of records.
* For each rule, we add a set of negative conditions ($NOT$ statements in the rule), removing non-relevant records in a group: a condition removing the largest number of non-relevant records is found, and then iteratively, more are added with the same logic, until no conditions are found that would not also remove positive records.
* The extended set of rules is sorted by the difference between the number of positive and negative records they identify in descending order. Given this ordered list, the cumulative number of unique positive records is computed and used to group the rules: each rule in the same group of rules adds the same number of positive records to the preceding group but are ordered according to the number of negative records they identify.
* These groups of rules are outputted, and the researcher is asked to review them, selecting one or more (useful if they convey radically different meaning) rules from each group, or edit them (in case too specific positive or negative conditions were included). By default, the framework suggests keeping the first rule of each group since it would identify the lower number of non-relevant records for the same number of positive one (.e., higher specificity). It is possible to exclude a group of rules altogether, especially the last ones with very low positive/negative records ratios.
* The selected rules are joined together by $OR$ statements, defining a subset of records with a sensibly higher proportion of positive records than the original one.
* Given this data set, the next step removes redundant rules (i.e., rules whose positive records are already included in more specific ones) and conditions (i.e., conditions that once removed do not decrease the total number of positive or do not increase the negative records).
* Finally, the rules are re-elaborated in a format usable on the major scientific databases.

It is important to note that this process is entirely data-driven. The algorithm is only aware of the "world" defined by the data set, itself generated by a specific search query focused on a particular topic. Therefore, the new query may not be specific once applied to an unbounded search domain, returning an unmanageable amount of unrelated results. We suggest extending the query by adding domain-specific terms linked by an $AND$ statement, separating the original query and the new addendum in two distinct logical blocks by enclosing them in parentheses. See the Results for the specific query used in the second session. <mark>*not sure whether to put this part here or in the discussion*</mark>

### Hyperparameter search
The machine learning model behind the algorithm has a number of hyperparameters in addition to the BART specific ones: the number of ensemble models, the source of randomness between models (just derived from MCMC sampling, the default, or whether to bootstrap the data first), the oversampling of positive records, the PrI quantiles, the number of initial manually labelled records. These parameters have a relevant effect on the algorithm performance.\
We set up a grid search to evaluate the parameters on a subset made of the first 1200 records from the first session; these records were fully labelled as relevant or not. We run the framework for each combination of parameter values, automatizing the manual review part using the already collected labels. The framework ran until four CR iterations with no positive records were returned, or the whole dataset was labelled.\
For each combination, a performance score was computed as the product of *Efficiency* (1 minus the number of records that required review over the total) and *Sensitivity* (number of positive records found over the total of positives). We then identified homogeneous "performance clusters" of parameter values using a decision tree on the score. For the final analysis, we chose the best cluster of parameter value combinations and, inside the cluster, the best combination in order of Sensitivity and Efficiency.

### Performance evaluation
It is hard to estimate the generalized performance of such a model since the data is not a random sample (the records are presumably the whole population of articles from the searched databases for those search criteria), and since the order of records matters (it drives the choice of the initial labelling set). Therefore we postulated that usual random train/test methods to estimate the out-of-sample machine learning performance [@kohavi1995study] would not be representative in this situation. Also, common time-series based predictive performance methodologies would reveal inadequate [@tashman2000out] since the expected proportion of positive matches is not constant throughout the data set and is expected to fall fast as the first matches are found.\
Our solution was to use Bayesian logistic regression to model the label of the manually reviewed records on the lower bound of the [98% PrI] produced by our algorithm. For the model we used weakly regularizing, robust priors for the intercept (Student T with $\nu=3,\mu=0,\sigma=2.5$) and the linear coefficient (Student T with $\nu=3,\mu=0,\sigma=1.5$). The goal of the model is to estimate the PPD of the missed number of positive matches in the whole data set. Given that this model is conditional only on the BART predictions and not on the DTM, it is characterized by more uncertainty, providing a plausible worst-case scenario. The quality of the model was evaluated through Bayesian $R^2$ [@gelman2019r] of which we reported the posterior median and 90% Credible Interval [90% CrI].\
The predictive distribution of the number of missed positive records allows to estimate the expected long-run *Sensitivity* and the *Work saved over random* (WSoR) of the algorithm. The WSoR is based on a negative hypergeometric model to estimate the number of records to manually label to find the same number of positives if records were evaluated in random order [@chae1993presenting]; the WSoR is then one minus the ratio of the reviewed records over this number. For the number of predicted positive records, the sensitivity and WSoR, we reported the truncated 90% PrI [trunc. 90% PrI], which is the uncertainty interval bounded at the number of observed total positive records: since each positive match is manually verified, the probability of a number of total positive records lower than observed is zero.

>>>>

An initial training set of manually labelled records is needed to start the automatic classification. We suggest manually label the first 250 records (see "hyperparameter optimization" later).\

### Document-Term matrix generation
The annotation file produced in the previous step has a number of fields characterizing a scientific publication. The framework models the relevance of a record based on the following fields: title, abstract, authors, keywords, MESH terms [@lipscomb2000medical]. A series of Natural Language Processing (NLP) techniques [@baeza1999modern; @marshall2015systematic; @ananiadou2006text] are employed to transform the textual information in these fields into features for machine learning, through a bag-of-words approach [@marshall2015systematic]. The processing of free text fields includes tokenization (i.e., extracting the terms), common stopwords (i.e. sentence components bringing no meaning) removal, part-of-speech filtering (only nouns, adjectives, verbs and untagged terms are retained), and lemmatization of the terms (i.e. reduction to their base grammar form). Processing for authors, keywords and MESH terms identify logical units (e.g., author's full names, composite keywords) without breaking them into single terms, and no stopwords removal or lemmatization are applied.\
To reduce noise and save computation time, terms that appeared in less than 5% of the labelled documents (positive and negatives) are removed from negative records. All terms in the positive set are kept to increase sensitivity at the cost of specificity.\
Some terms tend to co-appear in records (non-consecutive ngrams, nc-ngrams), often carrying a particular meaning when copresent. To detect nc-ngrams, we generated a network representation posing edges between terms with a cosine similarity in terms of records copresence > 0.5. We extracted the maximal cliques in the network [@eppstein2010listing] representing highly correlated groups of terms; These generated terms are added to the data set. To avoid overfitting, we kept nc-ngrams of a maximum of ten terms.\
A similarity network is built again using a similarity threshold of 0.9 and finding the cliques again. In this case, the cliques represent terms that always appear together and therefore can be considered synonyms. These terms are merged in the data set to increase computation efficiency and reduce overfitting.\
The final output is a matrix, also called a Document-Term Matrix (DTM), with $N_d$ rows representing the records $D$, $N_t$ terms column for the $t_{field}$ terms (divided by record field) and ${0,1}$ values whether $t_{field} \in D$. We also enriched the DTM with features storing the number of terms in each field to help the model scale term importance based on the document length.

@@

$$\pi_i = \frac{1}{M}\sum_{j=1}^MPr(L_i = \text{pos}|DTM,m_j)$$

as the PPD of a record $D_i$ being assigned a positive label ($L_i$), averaging the PPDs of the ensemble of $M=10$ models $m$, and

@@

* We fit a partition tree [@rpart] between the DTM and 800 samples from the PPD; if a term is present multiple times in the DTM (e.g. both title and abstract), they are counted just one, and field term count features are removed. This step generates a list of rules composed by $AND$/$NOT$ "conditions" made of terms/authors/keywords/MESH, which together identify a group of records.
* For each rule, we add a set of negative conditions ($NOT$ statements in the rule), removing non-relevant records in a group: a condition removing the largest number of non-relevant records is found, and then iteratively, more are added with the same logic, until no conditions are found that would not also remove positive records.
* The extended set of rules is sorted by the difference between the number of positive and negative records they identify in descending order. Given this ordered list, the cumulative number of unique positive records is computed and used to group the rules: each rule in the same group of rules adds the same number of positive records to the preceding group but are ordered according to the number of negative records they identify.
* These groups of rules are outputted, and the researcher is asked to review them, selecting one or more (useful if they convey radically different meaning) rules from each group, or edit them (in case too specific positive or negative conditions were included). By default, the framework suggests keeping the first rule of each group since it would identify the lower number of non-relevant records for the same number of positive one (.e., higher specificity). It is possible to exclude a group of rules altogether, especially the last ones with very low positive/negative records ratios.
* The selected rules are joined together by $OR$ statements, defining a subset of records with a sensibly higher proportion of positive records than the original one.
* Given this data set, the next step removes redundant rules (i.e., rules whose positive records are already included in more specific ones) and conditions (i.e., conditions that once removed do not decrease the total number of positive or do not increase the negative records).
* Finally, the rules are re-elaborated in a format usable on the major scientific databases.
Since the algorithm is data-driven and creates rules that are specific for the dataset but may be not once applied to comprehensive research databases, we added a new subquery in $AND$ specifying the general topics of our search.

### Hyperparameter search
The machine learning model behind the algorithm has a number of hyperparameters in addition to the BART specific ones: the number of ensemble models, the source of randomness between models (just derived from MCMC sampling, the default, or whether to bootstrap the data first), the oversampling of positive records, the PrI quantiles, the number of initial manually labelled records. These parameters have a relevant effect on the algorithm performance.\
We set up a grid search to evaluate the parameters on a subset made of the first 1200 records from the first session; these records were fully labelled as relevant or not. We run the framework for each combination of parameter values, automatizing the manual review part using the already collected labels. The framework ran until four CR iterations with no positive records were returned, or the whole dataset was labelled.\
For each combination, a performance score was computed as the product of *Efficiency* (1 minus the ratio of records that required review over the total) and *Sensitivity* (number of positive records found over the total of positives). We then identified homogeneous "performance clusters" of parameter values using a decision tree on the score. For the final analysis, we chose the best cluster of parameter value combinations and, inside the cluster, the best combination in order of Sensitivity and Efficiency.

### Performance evaluation
The citation data is not a random sample and it is ordered and for this reason, simple random sample based predictive performance approaches would not be useful (see Discussion). We focused on estimating the expected total sensitivity in the dataset, using a Bayesian logistic regression to model the label of the manually reviewed records on the lower bound of the [98% PrI] produced by our algorithm and then using this last model to produce a predictive cumulative distribution of the number of missed positive records. For this model, we used weakly regularizing, robust priors for the intercept (Student T with $\nu=3,\mu=0,\sigma=2.5$) and the linear coefficient (Student T with $\nu=3,\mu=0,\sigma=1.5$). Given that this model is conditional only on the BART predictions and not on the DTM, it is characterized by more uncertainty, providing a plausible worst-case scenario. The quality of the model was evaluated through Bayesian $R^2$ [@gelman2019r] of which we reported the posterior median and 90% Credible Interval [90% CrI].\
The predictive distribution of the number of missed positive records allows to estimate the expected long-run *Sensitivity* and the *Work saved over random* (WSoR) of the algorithm. The WSoR is based on a negative hypergeometric model to estimate the number of records to manually label to find the same number of positives if records were evaluated in random order [@chae1993presenting]; the WSoR is then one minus the ratio of the reviewed records over this number. For the number of predicted positive records, the sensitivity and WSoR, we reported the truncated 90% PrI [trunc. 90% PrI], which is the uncertainty interval bounded at the number of observed total positive records: since each positive match is manually verified, the probability of a number of total positive records lower than observed is zero.
